---
title: 'Lab 2 - R as a GIS'
author: "Bailey McNichol"
date: "10/10/2021"
output:
  pdf_document: default
  html_document: default
---


# Lab 02: R as a GIS

### Read the instructions COMPLETELY before starting the lab

This lab builds on many of the discussions and exercises from class, including previous labs.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. The code must be contained in either a .R script or a .Rmd markdown document. As I need to run your code, any data you use in the lab must be referenced using **relative path names**. Finally, answers to questions I pose in this document must also be in the repository at the time you submit your link to Canvas. They can be in a separate text file, or if you decide to use an RMarkdown document, you can answer them directly in the doc.

## Data

The data for this lab can be found in the ```./data/CBW/``` directory within the course GitHub repository. 

Spatial datasets:

1. Streams_Opened_by_Dam_Removal_2012_2017.shp

2. Dam_or_Other_Blockage_Removed_2012_2017.shp

3. County_Boundaries.shp

Non-spatial datasets

1. BMPreport2016_landbmps.csv


## Working with tabular data

```{r setup, echo=TRUE, message=TRUE}
# setup
library(tidyverse)
library(sf)
library(tmap)
library(stringr)
```


A "join" is a method to join multiple tables together using a matching "key" found in both datasets. For example:

```{r show a join, echo=TRUE, message=TRUE}

# table of names
t.names <- tibble(key = c(1, 2, 3), 
             name = c("Huey", "Dewey", "Louis"))

# table of scores
t.scores <- tibble(name = c("Louis", "Huey", "Dewey"),
                   grade = c(99, 45, 33))

# combined them
t.joined <- left_join(t.names, t.scores, by = "name")
t.joined
```

A "left join" finds starts with the table on the "left" and then finds matches in the table on the "right". See the documentation using ```?left_join``` for more details and for other types of joins. Sometimes the attributes you're using to join the tables won't have the same name, in which case the syntax is different:

```{r diff syntax, echo=TRUE, message=TRUE}

t.wonkyNames <- tibble(nombre = c("Dewey", "Louis", "Huey"),
                       x = rep(999),
                       favoriteFood = c("banana", "apple", "carrot"))

t.joined2 <- left_join(t.names, t.wonkyNames, by = c("name" = "nombre"))
t.joined2
```

## Let's take a look at some tabular data

This dataset includes a list of best management practices ("BMPs") to reduce nutrient and sediment pollution in the Chesapeake Bay Watershed.

```{r tabularData, echo=TRUE, message=TRUE}

bmps <- read_csv("../data/CBW/BMPreport2016_landbmps.csv")

glimpse(bmps)

```

Look at the attribute "GeographyName" - it's a character attribute that contains the counties' FIPS code, but also some ancillary explanatory data we need to get rid of. There are multiple ways of doing so, including some (very) fancy automated methods that detect patterns of numbers and characters. We're going to take a simpler approach and assume that all FIPS codes are only 5 characters long.

```{r trim a string, echo=TRUE, message=TRUE}

# edit the bmps variable in place, which isn't always best practices
bmps <- bmps %>% mutate(., FIPS.trimmed = stringr::str_sub(GeographyName, 1, 5))
```

This can be sueful when you're trying to create "keys" by which to join tables or just clean your tables in general

Let's recall how to do some simple tasks

```{r review, echo=TRUE, message=TRUE}

# Let's calculate the total cost by BMP and then plot it
bmps %>% group_by(BMPType) %>% summarise(totalCost = sum(Cost)) %>%
  ggplot(., aes(x = BMPType, y = totalCost)) +
  geom_bar(stat = "identity") +
  theme_minimal()

# Doesn't really work. This is because there are missing data in the cost attribute. Let's look at it (and this is why we do exploratory data analysis first)

summary(bmps$Cost)

# Yup, lots of "NA's"... We can drop them in our analysis. Look carefully at the code in the 'sum' function


bmps %>% group_by(BMPType) %>% summarise(totalCost = sum(Cost, na.rm = T)) %>%
  ggplot(., aes(x = BMPType, y = totalCost)) +
  geom_bar(stat = "identity") +
  theme_minimal()

```

We can also group by multiple variables at the same time. For example:

```{r review2, echo=TRUE, message=TRUE}
# group by state and sector, sum total cost
twofactors <- bmps %>% group_by(StateAbbreviation, Sector) %>% summarise(totalCost = sum(Cost))
twofactors
```

For our last bit of review, let's make a few box plots:

```{r review3, echo=TRUE, message=TRUE}

# A simple one
bmps %>% ggplot(., aes(x = StateAbbreviation, y = AmountCredited)) +
  geom_boxplot(aes(fill = StateAbbreviation))

# Very heavily skewed, so just for the sake of visualization, let's subset the data (dramatically)
bmps %>% 
  dplyr::filter(., AmountCredited > 1 & AmountCredited < 100) %>% 
  ggplot(., aes(x = StateAbbreviation, y = AmountCredited)) +
  geom_boxplot(aes(fill = StateAbbreviation))

# We can also plot multiple dimensions in our plot using the `facet` family of commands in ggplot

bmps %>% 
  dplyr::filter(., AmountCredited > 1 & AmountCredited < 100) %>% 
  ggplot(., aes(x = StateAbbreviation, y = AmountCredited)) +
  geom_boxplot(aes(fill = StateAbbreviation)) +
  facet_grid(Sector~.)

```

The last new item uses the ```%in%``` command. It's a way to quickly figure out which elements are inside of another. In that sense, it's similar to a spatial intersection, but for other types of data.

```{r using in, echo=TRUE, message=TRUE}

x <- c(1, 2, 3, 4, 5)

# is 7 in our vector?
7 %in% x # should be False

2 %in% x # should be True

# can also do it with vectors
c(4, 99, 1) %in% x
```


Lastly, let's recall using tmap on our data. Also remember you can use ```sf::st_make_valid``` to fix offending geometry

```{r remembering tmap, echo=TRUE, message=TRUE}

counties <- sf::read_sf("../data/CBW/County_Boundaries.shp")
counties %>% sf::st_is_valid()
counties <- counties %>% sf::st_make_valid()

# quick map of the data
tm_shape(counties) + tm_polygons(col = "ALAND10")
```

\newpage

## Your tasks

Using the following data...

```{r tasksetup, echo=TRUE, message=TRUE}

# spatial
counties <- sf::read_sf("../data/CBW/County_Boundaries.shp") %>% sf::st_make_valid()
dams <- sf::read_sf("../data/CBW/Dam_or_Other_Blockage_Removed_2012_2017.shp") %>% sf::st_make_valid()
streams <- sf::read_sf("../data/CBW/Streams_Opened_by_Dam_Removal_2012_2017.shp") %>% sf::st_make_valid()

# aspatial
bmps <- read_csv("../data/CBW/BMPreport2016_landbmps.csv")
```

...complete the following tasks.

Complete each task COMPLETELY USING R CODE. YOU MUST SHOW YOUR WORK FOR EACH ANSWER. Label your variables sensibly and use comments such that I can find your answers and your work. The following tasks draw upon lecture, your assigned readings, and the examples shown above. As always, there are multiple ways of completing each task. Remember, it's always a good idea to peform some exploratory data analysis on your own prior to starting work.

## Lab 2 Tasks

### Task 1: Basic data manipulation

#### 1.1 Calculate summary statistics for the Cost of BMPs for each State (including DC)

First, I grouped the BMPs by state, and then used summarize to calculate the mean, median, minimum, and maximum cost of BMPs for each State. The output is this table of values, where each row is a state.
```{r include=TRUE, echo=TRUE}
# Look at the variable names
glimpse(bmps)

# Use group_by to group by states
bmps %>% group_by(StateAbbreviation) %>% 
  # Use summarize to calculate the summary statistics
             summarise(meanCost = mean(Cost, na.rm = T),
                       medCost = median(Cost, na.rm = T),
                       minCost = min(Cost, na.rm = T),
                       maxCost = max(Cost, na.rm = T))

```


#### 1.2 Make a scatterplot of Cost vs. TotalAmountCredited, ONLY FOR Units of type "Acres". You may need to apply a data transformation to one or more axes if the data are heavily skewed.

First, I filtered the BMPs to only include observations where the units were acres, and then I dropped all NA values for Cost. Then, I plotted the Cost of BMPs versus the Total Amount of Credited Acres. To improve data visualization and deal with the heavy left-skew of the cost data (losts of small values with a few large outliers), I applied a log base 10 transformation to Cost, first adding 0.0001 to each value to accommodate values of 0 for Cost (and we can see that there are many different total amount credited values that had a cost of $0 - the weird vertical line of data on the left). I chose to leave in the zero values on my figure, as there are many credited acres that a cost of zero (so I didn't want to exclude relevant data). On the x-axis of the figure, the log 10 values can be back-transformed to represent costs of approximately 0.003, 1.00, 316.3, 100,023, and 31,630,060 dollars, respectively.
```{r include=TRUE, echo=TRUE}

bmps %>% 
  # Only include units of acres
  dplyr::filter(., Unit == "Acres") %>% 
  # Remove NAs in Cost
  drop_na(Cost) %>% 
  # Transform Cost to log10 scale, adding 0.0001 to each value to avoid undefined numbers
  ggplot(., aes(x = log10(Cost+0.0001), y = TotalAmountCredited)) +
  geom_point() +
  theme_bw() +
  labs(x = "Log(Cost of BMPs) ($)", y = "Total Amount Credited (Acres)") 

# Back-transform the log10 values to actual costs in dollars
10^c(-2.5001, 0.0001, 2.5001, 5.0001, 7.5001)
```


#### 1.3 Make a boxplot with "StateAbbreviation" on the x-axis and "TotalAmountCredited" on the y-axis. HOWEVER, the only data I want plotted are for cover crop BMPs. Note, there are many types of cover crops in this dataset, and I want you to include them ALL. There are handy functions within the ```stringr``` package that can help you here.

First, I filtered the data to include any BMPShortName values with the string "covercrop" using the filter and str_detect functions. Then, I was able to produce the boxplot of Total Amount Credited by State; I opted to leave the outliers on the figure, as there are many and there is no clear cut-off that improves the data visualization considerably.
```{r include=TRUE, echo=TRUE}
bmps %>%
  # Use str_detect within filter to select only (and all of) the cover crop BMPs
  filter(str_detect(BMPShortName, "covercrop")) %>% 
  # Plot the variables
  ggplot(., aes(x = StateAbbreviation, y = TotalAmountCredited)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "State", y = "Total Amount Credited") 

```


#### 1.4 make a scatterplot of the dam dataset, this time with "YEAR" on the x-axis and "STATE" on y-axis (think of it like a timeline). Assume no dams were built in year 0, so you'll need to remove those data points.

First, I filtered the YEAR variable to exclude values of zero, which eliminated MD (all YEAR values were zero for Maryland). Then, I removed NA values for YEAR, and made a scatterplot of the Year of Dam Establishment (YEAR) by STATE - the data now only includes PA and VA, and only data on one dam for VA.
```{r include=TRUE, echo=TRUE}

dams %>%
  # Filter to remove 0 values for YEAR
  filter(YEAR != 0) %>% 
  # Remove NAs for year
  drop_na(YEAR) %>% 
  # Plot the variables
  ggplot(., aes(x = YEAR, y = STATE)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year of Dam Establishment", y = "State") 

```


#### 1.5 make one last (aspatial) visualization. But this time, it's your choice what data and plots to use. The only requirement is that you link two of the datasets together in some manner. Be creative. Make it look nice (e.g., use proper labels, interesting colors/shading/size).

I created a new variable in the bmps dataset that represents the County FIPS Code, following the example in the introduction. Then, I joined the counties and bmps datasets using the County FIPS codes (called GEOID10 in counties). I assessed the levels (unique values) of the BMPs, and then filtered this new dataset to only include BMPs involving wet areas. Then, I created a figure showing the relationship between water area and numbers of different BMPs that rely on water, filtering the data for visualization to exclude outliers and be easier to interpret (changing label angles, etc).
```{r include=TRUE, echo=TRUE}

# Create County FIPS Codes for bmps data
bmps <- bmps %>% mutate(., FIPS.trimmed = stringr::str_sub(GeographyName, 1, 5))

# Join the bmps data with the counties, by county code
county_bmp <- left_join(counties, bmps, by = c("GEOID10" = "FIPS.trimmed"))

# Examine the levels of BMPShortName
levels(as.factor(county_bmp$BMPShortName))

# Plot the number of BMPs involving wet areas versus the total area of water
county_bmp %>%
  # Filter BMPShortName to only include BMPs in wet areas
    filter(str_detect(BMPShortName, "wet")) %>% 
  # Remove outlier values of Water Area for visualization
    filter(AWATER10 < 150000000) %>% 
  # Create a boxplot of water area across BMPs involving wet areas
    ggplot(., aes(x = BMPShortName, y = AWATER10)) +
  # Change fill color to orchid
    geom_boxplot(fill = "orchid") +
    theme_bw() +
  # Make x-axis labels readable
    theme(axis.text.x=element_text(colour="black", 
                                 angle = 40, hjust = 1)) +
    labs(x = "Best Management Practice", y = "Water Area (m2)", 
         title = "Relationship between Numbers of Best Management Practices \n 
         Involving Wet Areas and Water Area") 

```


### Task 2: Spatial operations

#### 2.1 Find the 5 longest streams in the 'streams opened by dam removal' dataset

First, I arranged the rows of data from longest to shortest stream length (LengthKM). Then, I used the slice function to just subset first five rows of data. The five longest streams in descending order are: Little Chiques Creek, Toms Run, Johns Run, Little Trough Creek, and Great Trough Creek.
```{r include=TRUE, echo=TRUE}

streams %>% 
    # Arrange the data by stream length, in descending order
    arrange(desc(LengthKM)) %>% 
    # Subset the data with slice to just include the top 5 rows
    slice(1:5)
```

#### 2.2 Find the three counties with the greatest TOTAL length of streams (opened by dam removal) in them

First, I joined the counties and streams spatial data using st_join with an st_intersects join. Then, using this new joined data, I grouped values by county, removed NAs for stream length, calculated the total stream length per county with summarize, reordered these values to be in descending order, and then selected the three counties with the greatest total length of streams. The top three counties were: Augusta, Huntingdon, and Cameron.
```{r include=TRUE, echo=TRUE}

# Spatially join the counties and streams data using st_intersects
co_streams <- st_join(counties, streams, join = st_intersects)

co_streams %>%
  # Group the joined data by county name
    group_by(NAME10) %>%
  # Remove NA values for stream length
    drop_na(LengthKM) %>%
  # Create a variable that represents the sum (total) length of streams in each county
    summarize(totalStreamsKM = sum(LengthKM)) %>%
  # Arrange the total length of streams per county in descending order
    arrange(desc(totalStreamsKM)) %>% 
  # Show the top 3 counties
    slice(1:3)
  
```

#### 2.3 Make a map of the counties, shading each county by the total cost of BMPs funded/implemented in that county. This will required you to join multiple datasets together

First, I used the same code as above to trim and create a county FIPS code in the bmps data using str_sub(). Then, I grouped the bmps data by county, removed NA values for the Cost variable, and calculated the total cost of BMPs by county using summarise(), creating a new variable, totalCost, in a new tibble, BMPs. I then performed a left join by county on the counties and BMPs data sets, and mapped the counties, color coded by cost. We see on the map that most counties had a total cost of BMPs less than or equal to $50 million, although a few counties had higher costs.
```{r include=TRUE, echo=TRUE}


BMPs <- bmps %>% 
     # Create County FIPS Codes for bmps data with str_sub()
        mutate(., FIPS.trimmed = stringr::str_sub(GeographyName, 1, 5))  %>%
     # Group the BMP data by county
        group_by(FIPS.trimmed) %>%
     # Remove NA values for Cost
        drop_na(Cost) %>%
     # Calculate the total cost of BMPs in that county
        summarise(totalCost = sum(Cost))
        
# Join the bmps data with the counties, by county code
county_bmp <- left_join(counties, BMPs, by = c("GEOID10" = "FIPS.trimmed"))

# Map counties and shade by BMP cost
tm_shape(county_bmp) +
          tm_polygons(col = "totalCost") +
          tm_layout(legend.title.size = 1.5,
          legend.text.size = 0.6,
          legend.position = c("left","top"))

```


#### 2.4 For each removed dam, find the closest stream segment

First, I reduced the sizes of the dams and streams datasets by only selecting the relevant varibles in each with select(). Then, I used st_nearest_feature() to find the nearest stream to each dam, and then took it a step further by calculating the distance between the dam/stream (not surprisingly, most were very close). Finally, I used cbind() to combine the dam names, nearest stream names, and information about the distance between them and the geometry of each feature. I have also included an alternative method that automatically calculates the nearest feature and performs the join in one join, with st_join() and method st_nearest_feature().
```{r include=TRUE, echo=TRUE}

# Create a new tibble with only the relevant variables in the dams data
damsGeom <- dams %>% 
          dplyr::select(STATE,DAM_NAME, geometry) 
# Create a new tibble with only the relevant variables in the streams data
streamsGeom <- streams %>% 
          dplyr::select(GNIS_Name,LengthKM,Shape_Leng, Shape_Le_1, geometry) 

# Determine the closest stream segment to each dam using st_nearest_feature()
(nearest = st_nearest_feature(damsGeom,streamsGeom))
# Calculate the distance between each dam and the closest stream
(dist = st_distance(damsGeom,streamsGeom[nearest,], by_element=TRUE))
# Create a tibble that includes the dam names, nearest stream names, and distances between them
(DamStreamjoin = cbind(damsGeom, streamsGeom[nearest,], dist))


## Alternative method that automatically joins the data
st_join(damsGeom, streamsGeom, st_nearest_feature)

```


#### 2.5 Calculate how many removed dams are (or were) in each state

First, I grouped the dams data by state using group_by(), and then used the n_distinct() function within summarize() to count the number of removed dams per state (all dams in this dataset were removed, so no extra filtering was required). We see that 2 dams were removed in Maryland, 27 were removed in Pennsylvania, and 5 were removed in Virginia.
```{r include=TRUE, echo=TRUE}

dams %>% 
     # Group the BMP data by county
        group_by(STATE) %>%
     # Calculate the total cost of BMPs in that county
        summarise(NumberDamsRemoved = n_distinct(DAM_NAME))
```



